Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-06 paper-ieee-trans.tex
Extend CI implementation and figure

The software quality policies \cite{indigo-d31} cover: 1) the identification
and description of the \emph{quality requirements} that the software need to
comply with, and 2) the \emph{quality metrics}.

The requirements are checked at early stages in the software development
process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so that any
bug or design issue is likely to be detected and corrected promptly. The source
code is publicly available in GitHub repositories under the project organization called
{\sl indigo-dc} \cite{indigo-github} to increase the visibility of the products
catalogue, promoting the external contributions and software adoption. Furthermore,
the source code is checked for compliance with community de--facto style standards,
selected for each one of the variety of programming languages being used as seen in
Figure \ref{fig:fig_codestyle}.

@@

The DevOps methods emphasize on exercising SQA techniques to avoid infrastructure
disruption whenever new developments are deployed into production systems. The 
following DevOps approaches have been progressively adopted throughout the project's
lifetime, being one of its major outcomes.

\subsubsection{Continuous Integration}
The INDIGO--DataCloud project promoted the application of a CI scenario to enforce
the SQA requirements -- described in Section \ref{subsec:sqa} -- for any piece of
software being produced within the project. Such environment requires an automation
ready--to--go infrastructure where the different technologies involved -- source code
management platform, automation server and containerization -- interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is comprised of diverse quality checks, enforcing at least the successful
execution of code style and minimum unit testing coverage. To take over the 
implementation of such infrastructure, the project leveraged on tightly integrated open
source tools such as GitHub \cite{github}, as the online source code repository,
and Jenkins \cite{jenkins}, the event-response CI system. Docker container provisioning
is used to provide instant computing power needed for executing the pipeline jobs.

This CI approach is guiding the project’s software development phase throughout the
first and second major releases.

>>>>

The software quality policies \cite{indigo-d31} cover: 1) the identification
and description of the \emph{SQA requirements} that the produced software need 
to comply with, and 2) the \emph{quality metrics}, to monitor each software product's 
behaviour throughout the development, release and production stages.

\subsubsection{SQA criteria}

The SQA requirements cover the set of conventions and recommendations that pave the way for
an adequate development, timely deliver and reliable operation of the software components
produced. The SQA criteria is in continuous evolution, reviewed periodically to match the
project's needs.

\begin{itemize}
\item \textbf{Code style}.
Code style standards are enforced for every software component in the project stack. Community
most-adopted or \textit{de-facto standard} is initially recommended, however the developer teams eventually
choose the set of guidelines to comply with. The ultimate goal for code style assessment is to
improve the readability and reusability of the source code produced under the scope of the
project.

\item \textbf{Unit and Functional testing}.
Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
testing} in this context is done enforcing the successful execution of the previous
functionality tests already present in the code, so to uncover any malfunction introduced by
the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
focusing on the code's internal design. Project's SQA criteria sets a recommended threshold of
70\% coverage for unit testing.

\item \textbf{Code review}.
Represents the last step in the change management pipeline, once the candidate change has 
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequance in terms of scope, objective fulfillment, 
documentation completeness, etc. On approval, the candidate change is definitely merged into
production.

\item \textbf{Documentation and Automated Deployment}.
Last but not least, the SQA criteria sets the path for the adoption of the developed software by
defining the documentation content, according to the target audience. These requirements also 
promote the automation, both in terms of documentation creation and service deployment. The
\textit{documentation is treated as code} so that it is portable and easily reviewed, using the same workflow
as the code does, validating the changes before being updated into the production repository. On the
other hand, the software component deployment is requested to be automated using a configuration 
management tool. These tools allow to have an \textit{infrastructure as code}, managing the component's
deployment through declarative definitions. The definition files appear as self-documenting 
code, as they sequentially guide the component's deployment process in multiple platforms.
\end{itemize}

%integrated in the Jenkins CI service \cite{indigo-jenkins}, so that any
%bug or design issue is likely to be detected and corrected promptly. The source
%code is publicly available in GitHub repositories under the project organization called
%{\sl indigo-dc} \cite{indigo-github} to increase the visibility of the products
%catalogue, promoting the external contributions and software adoption. Furthermore,
%the source code is checked for compliance with community de--facto style standards,
%selected for each one of the variety of programming languages being used as seen in
%Figure \ref{fig:fig_codestyle}.

@@

The DevOps methods emphasize on exercising SQA techniques to avoid infrastructure
disruption whenever new developments are deployed into production systems. The 
following DevOps approaches have been progressively adopted throughout the project's
lifetime, being one of its major outcomes in terms of sofware reliability.

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth, height=50mm]{images/jenkins_CI_builds.png}
\caption{.}
\label{fig:fig_jenkins_CI_builds}
\end{figure}

\subsubsection{Continuous Integration}
\label{subsec:ci}
The INDIGO--DataCloud project promoted the application of a CI scenario that enforces,
for any piece of software produced, the testing requirements defined in the SQA criteria,
as described in Section \ref{subsec:sqa}. Such environment requires an automation
ready--to--go infrastructure where the different technologies involved -- source code
management platform, automation server and containerization -- interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is mainly comprised by the code style validation, unit and functional testing
coverage. The CI pipeline is complemented with additional quality checks, such as 
integration tests for the software components where automation of this type of testing
is applicable, or security linters for the static analysis of suspicious
constructs that could lead to security risks. Metrics gathering also takes part of
the CI pipeline so to have a per-change trend evolution of the common source code 
related metrics, such as the source lines of code (SLOC) or the cyclomatic complexity.

The practical implementation of the CI scenario leveraged on tightly integrated open
source tools such as GitHub \cite{github} -- as the online source code repository --,
Jenkins \cite{jenkins} -- the event-response CI system -- and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs. 
The project defined a source code contribution workflow, based on GitHub Pull Requests 
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline report their exit status back to GitHub, preventing the change to be
approved if they are not successfully executed.

Figure \ref{fig:fig_jenkins_CI_builds}

The described CI approach is guiding the project’s software development phase throughout
the first and second major releases. Throughout its lifetime, 
% FIXME complement with the paragraph above and the references to the figures for code style and unit testing!!!!
%The implementation of a CI scenario is accomplished by automating the execution of
%the test types defined in the SQA criteria described in Section \ref{subsec:sqa}. 
%Each change in the source code will trigger the associated with the software component.
%As each change is validated, the chances of early detection of defects increases. Within
%this scenario, the cost of defect solving is dramatically reduced and the reliability of the 
%software solutions improved, as any bug or design issue is likely to be detected and subsequently
%corrected at this phase.
%
